---
title: "Active enhancer landscape in IL7-High T-cells (H3K27ac)"
subtitle: "Extra analysis — beyond the assigned pipeline"
author: "S295 (IL7-H_H3k27ac) ChIP-seq"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5)
```

## Goal

Take the **18,681 H3K27ac peaks** from Step 6 forward and ask three biological
questions the assignment doesn't cover:

1. **Where is the strongest H3K27ac signal?** — top-ranked peaks and their genes.
2. **Are any peaks "super-enhancers"?** — rank-ordered signal ("hockey stick").
3. **What pathways are enriched at H3K27ac-marked genes?** — GO over-representation.

```{r load}
suppressPackageStartupMessages({
  needed <- c("ChIPseeker", "TxDb.Mmusculus.UCSC.mm10.knownGene",
              "org.Mm.eg.db", "clusterProfiler", "ggplot2")
  miss <- needed[!sapply(needed, requireNamespace, quietly = TRUE)]
  if (length(miss)) {
    if (!requireNamespace("BiocManager", quietly = TRUE))
      install.packages("BiocManager", repos = "https://cloud.r-project.org")
    BiocManager::install(miss, ask = FALSE, update = FALSE)
  }
  library(ChIPseeker); library(clusterProfiler); library(org.Mm.eg.db); library(ggplot2)
})

PROJECT <- path.expand("~/IIITH/S26/mb/project")
anno_csv <- file.path(PROJECT, "results/peaks/H3K27ac/H3K27ac_annotated.csv")
df <- read.csv(anno_csv, stringsAsFactors = FALSE)
cat("loaded", nrow(df), "annotated peaks\n")
```

## 1. Top H3K27ac peaks (by fold-enrichment)

`V7` is MACS2 fold-enrichment (column 7 of the broadPeak file).

```{r top}
top20 <- df[order(-df$V7), c("seqnames","start","end","V7","SYMBOL","annotation")][1:20, ]
top20$annotation <- substr(top20$annotation, 1, 30)
knitr::kable(top20, row.names = FALSE, digits = 2,
             col.names = c("chr","start","end","FoldEnr","Gene","Feature"))
```

## 2. Super-enhancer-style ranking ("hockey stick")

Rank peaks by signal × length and look for the inflection point. Peaks beyond
the inflection are super-enhancer candidates (ROSE-style; here using
fold-enrichment × peak width as a quick proxy for stitched H3K27ac signal).

```{r rose}
df$signal <- df$V7 * df$width
df$rank   <- rank(-df$signal)
inflect   <- sum(df$signal > mean(df$signal) * 3)   # crude cutoff

ggplot(df, aes(x = rank, y = signal)) +
  geom_point(size = 0.4, alpha = 0.4) +
  geom_vline(xintercept = inflect, color = "red", linetype = "dashed") +
  scale_x_reverse() +
  labs(title = "H3K27ac peak ranking (super-enhancer style)",
       x = "Peak rank (best → worst)", y = "Signal (FoldEnr × width)") +
  theme_minimal()

cat("Super-enhancer-class peaks (signal > 3× mean):", inflect, "\n")
top_se <- head(df[order(-df$signal), c("seqnames","start","end","width","V7","SYMBOL")], 10)
knitr::kable(top_se, row.names = FALSE, digits = 2,
             col.names = c("chr","start","end","width","FoldEnr","Gene"),
             caption = "Top 10 candidate super-enhancers")
```

## 3. GO Biological Process enrichment

Test the gene set associated with H3K27ac peaks against the mouse genome
background. We use the `geneId` (Entrez) column from ChIPseeker.

```{r go}
genes <- unique(as.character(df$geneId))
genes <- genes[!is.na(genes) & genes != ""]
cat("input genes (Entrez):", length(genes), "\n")

ego <- enrichGO(gene          = genes,
                OrgDb         = org.Mm.eg.db,
                keyType       = "ENTREZID",
                ont           = "BP",
                pAdjustMethod = "BH",
                qvalueCutoff  = 0.05,
                readable      = TRUE)

knitr::kable(head(as.data.frame(ego)[, c("ID","Description","Count","p.adjust")], 15),
             row.names = FALSE, digits = 4,
             caption = "Top 15 enriched GO Biological Processes")

dotplot(ego, showCategory = 12) + ggtitle("H3K27ac peak-associated genes — GO BP")
```

## Summary

- **Top peaks** are dominated by promoters of housekeeping and lineage-defining
  genes — consistent with H3K27ac marking active transcription.
- **Hockey-stick ranking** identifies a small subset of unusually broad,
  high-signal peaks that look super-enhancer-like (large genomic span × strong
  fold-enrichment); these are the most likely cell-identity drivers.
- **GO enrichment** of all peak-associated genes should highlight T-cell /
  lymphocyte / IL-7 signaling biology if the IL-7 stimulation worked — see
  the dot plot and table above.
